\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-17 1-intro.Rmd
more refine on intro

Spatio-temporal data concerns variables in the spatial coordinates across time. Some variables are invariant to time and can be solely identified by its spatial location while other time varying variables need both the spatial and temporal identifiers to identify. Manipulating data in these two levels is usually done separately and the third tidy data principal [@tidydata] recommends each type of observational unit to form a table. Another feature for a spatio-temporal data structure to have is that variables created from subsequent operations can be automatically added to the existing table with the same level. This is better than an output that index the variables with the identifiers since otherwise analysts will need to manually join this output with existing variables in the same level.

@@

Currently, available spatio-temporal data structures in R includes: \pkg{spacetime} [@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). \pkg{spacetime} is built on \pkg{sp} [@sp] and \pkg{xts} [@xts], both of which have been replaced by more recent implementations. \pkg{spatstat} [@spatstat] implements a `ppp` class for point pattern data; and more recent, \pkg{stars} [@stars] implements a spatio-temporal for vector and raster spatio-temporal data.  While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables at each temporal unit; or split them into two separate tables that analysts need to manually join the data constantly when doing the analysis. This creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any variables that are measured at fixed location across time.

>>>>

Spatio-temporal data concerns variables observed in the space across time and this paper focuses on the spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube and here a cube with three axes: spatial identifier, temporal identifier, and variable can be used to conceptualise spatio-temporal data. In this cubic formation, space only occupy one dimension, rather than two, to avoid the needs of hypercube for multivariate spatio-temporal data.

While conceptually the data format is relatively simple, the actual data object that comes to the analysts in the analysis can be in various forms. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) combines the time invariant and varying variables into a single table through duplicating the time invariant variables; 2) splits those variables into a separate spatial table and a temporal table; 3) ...; and lastly 4) is array-based data, like NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data format: (1) a single table for all the variables; (2) separate tables for spatial and temporal variables; (3) a table of metadata used to query the database and a separate table for queried data; and (4) an array data structure.", fig.align='center', out.height = "40%"}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

<!-- Manipulating data in these two levels is usually done separately and the third tidy data principal [@tidydata] recommends each type of observational unit to form a table. Another feature for a spatio-temporal data structure to have is that variables created from subsequent operations can be automatically added to the existing table with the same level. This is better than an output that index the variables with the identifiers since otherwise analysts will need to manually join this output with existing variables in the same level. -->

@@

Different authors have proposed various data objects to work with spatial temporal data in \proglang{R} and this includes: \pkg{spacetime} [@spacetime] is built on the \pkg{sp} [@sp] and \pkg{xts} [@xts] class and both of which have been replaced by more recent spatial and temporal data objects. \pkg{spatstat} [@spatstat] implements a `ppp` class for analysing point pattern data; and more recent, \pkg{stars} [@stars] has implemented an spatio-temporal array for wrangling both vector and raster data.

\textcolor{red}{Think again on the motivation and why it is hard to work existing tools.} In spatio-temporal data analysis, analysts will receive one of the various data formats depends on the data provider and need to manually rearrange the data to fit into one of the data object in \proglang{R} before starting any analysis on the data itself.

Looking at spatial or temporal data analysis, data objects like \pkg{sf} [@sf] and \pkg{tsibble} [@tsibble] have smoothed the data analysis in these two domains while there has not yet been a spatio-temporal data structure that makes it easy to explore spatio-temporal data and this creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any variables that are measured at fixed location across time.

<!-- While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables at each temporal unit; or split them into two separate tables that analysts need to manually join the data constantly when doing the analysis. \textcolor{red}{There has been data structure proposed for spatial (\pkg{sf} by @sf) and temporal (\pkg{tsibble} by @tsibble) data in the R community, while less has been done for spatio-temporal data.}  -->

### 2022-02-17 main.tex
more refine on intro

Spatio-temporal data concerns variables in the space across time.
\textcolor{red}{Elaborate on variable, space, and time}. If the time
interval is regular and each site has fixed point location, it can also
be thought of as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubical layout, variables can
be those that are invariant to time, i.e.~the longitude and latitude
coordinates, or those that varying across time. While conceptually the
data format is relatively simple, the actual data object that comes to
the analysts can be in various forms in practice. Figure
\ref{fig:illu-input} shows four examples where the first one combines
the time invariant and varying variables into a single table through
duplicating the time invariant variables; 2) splits those variables into
two separate sheets; 3) \ldots; and lastly 4) considers the array-based
data, like NetCDF (Network Common Data Form) data, which is commonly
used in earth science, e.g.~climatology, meteorology, and oceanography,
among others, to organise multivariate data in a spatio-temporal grid.

\textcolor{red}{think through a cube of long x lat x time vs. id x time x variable}

@@

\caption[An illustration of different spatio-temporal data format]{An illustration of different spatio-temporal data format: (1) a single table for all the variables; (2) separate tables for spatial and temporal variables; (3) a table of metadata used to query the database and a separate table for queried data; and (4) an array data structure of cube or hypercube.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

\textcolor{red}{more beginning like "working with such data different authors ..."}
Currently, there's various different spatio-temporal data structures in
R includes: \pkg{spacetime} \citep{spacetime}, which proposes four
space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI),
and trajectory (STT). \pkg{spacetime} is built on \pkg{sp} \citep{sp}
and \pkg{xts} \citep{xts}, both of which have been replaced by more
recent implementations. \pkg{spatstat} \citep{spatstat} implements a
\texttt{ppp} class for point pattern data; and more recent, \pkg{stars}
\citep{stars} implements a spatio-temporal for vector and raster
spatio-temporal data.

In spatio-temporal data analysis, analysts will receive one of the
various data formats depends on the data provider and need to manually
rearrange the data to fit into one of the data object in \proglang{R}
before starting any analysis on the data itself. Looking at spatial or
temporal data analysis, data objects like \pkg{sf} \citep{sf} and
\pkg{tsibble} \citep{tsibble} have smoothed the data analysis in these
two domains while there has not yet been a spatio-temporal data
structure that makes it easy to explore spatio-temporal data and this
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: climate observations from weather stations, air quality
variables from air monitoring stations, or any variables that are
measured at fixed location across time.

>>>>

Spatio-temporal data concerns variables observed in the space across
time and this paper focuses on the spatio-temporal vector data with
fixed locations and regular time interval. This type of data can
naturally been considered as a data cube and here a cube with three
axes: spatial identifier, temporal identifier, and variable can be used
to conceptualise spatio-temporal data. In this cubic formation, space
only occupy one dimension, rather than two, to avoid the needs of
hypercube for multivariate spatio-temporal data.

While conceptually the data format is relatively simple, the actual data
object that comes to the analysts in the analysis can be in various
forms. Figure \ref{fig:illu-input} shows four examples of wild
spatio-temporal data: 1) combines the time invariant and varying
variables into a single table through duplicating the time invariant
variables; 2) splits those variables into a separate spatial table and a
temporal table; 3) \ldots; and lastly 4) is array-based data, like
NetCDF (Network Common Data Form) data, which is commonly used in earth
science, e.g.~climatology, meteorology, and oceanography, among others,
to organise multivariate data in a spatio-temporal grid.

@@

\caption[An illustration of different spatio-temporal data format]{An illustration of different spatio-temporal data format: (1) a single table for all the variables; (2) separate tables for spatial and temporal variables; (3) a table of metadata used to query the database and a separate table for queried data; and (4) an array data structure.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

Different authors have proposed various data objects to work with
spatial temporal data in \proglang{R} and this includes: \pkg{spacetime}
\citep{spacetime} is built on the \pkg{sp} \citep{sp} and \pkg{xts}
\citep{xts} class and both of which have been replaced by more recent
spatial and temporal data objects. \pkg{spatstat} \citep{spatstat}
implements a \texttt{ppp} class for analysing point pattern data; and
more recent, \pkg{stars} \citep{stars} has implemented an
spatio-temporal array for wrangling both vector and raster data.

\textcolor{red}{Think again on the motivation and why it is hard to work existing tools.}
In spatio-temporal data analysis, analysts will receive one of the
various data formats depends on the data provider and need to manually
rearrange the data to fit into one of the data object in \proglang{R}
before starting any analysis on the data itself.

Looking at spatial or temporal data analysis, data objects like \pkg{sf}
\citep{sf} and \pkg{tsibble} \citep{tsibble} have smoothed the data
analysis in these two domains while there has not yet been a
spatio-temporal data structure that makes it easy to explore
spatio-temporal data and this creates a gap in the software development.
The requirement for such a tool is important given the ubiquity of
spatio-temporal vector data in the wild: climate observations from
weather stations, air quality variables from air monitoring stations, or
any variables that are measured at fixed location across time.
